something about variable-byte encoding

### 2020-02-19 gradu/msc_english_contents.tex
Added pseudocodes

\begin{table}
\centering
\caption{Size of support structure with SDSL-lite library}
\begin{tabular}{l||c c c c c c} 
Structure & TODO & get & multiple & size & calculations & here\\ 
\hline \hline 
$rank$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$select$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
\hline
%
\end{tabular}
\label{supportsize}
\end{table}

\section{DAC via rank}
Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. Their solution was to divide chunks in encoded bytes to separate arrays $A_1$ to 
$A_n$ and then use $rank_1$ to get direct access. For example, if an element needed 3 chunks when encoded, the least significant bits would go to $A_1$, second 
least significant bits to $A_2$ and the most significant bits would go to $A_3$. When fetching i-th element, getting the first chunk is just fetching $A_1$[i]. 
If the continuation bit is set at 0 (meaning there are more chunks to this element), getting the correct index for $A_2$ is calculated with $rank_1$($A_1$, i). 
Small values benefit from this reordering, because the first chunk does not need any calculation. This method causes a small overhead on the data, because each 
continuation bit array needs a support data structure for rank.

do rank/select algorithms + data structures need to be explained better? (or just referenced to)

<introduce Bri09 better?>

<example of Bri09>

@@

The blocks should be stored in an order that, depending on the endianness of the system, when reading an unsigned integer from the block array, the blocks are already 
in correct order. This way two bit shift operations are enough to extract the desired value.  Assume original element size is 32 bits, block length is b and requested
element is at position x in the block array and it's compressed within k blocks. Starting byte s of element e is x * b div 8 where div is the integer division. Offset o
is the remainder of previous division, o = x*b mod 8. 32-bit value e is read from memory from s. Then we bit shift left 32-b*k-o to remove trailing bits and bit shift right
32-b*k to re-align bits.

Simple way to calculate block length k of i-th element is to get $select_1$(i+1) and subtract previously calculated start block index from it. This however causes a second 
$select_1$ query, which costs a lot timewise. A much faster option is to iterate forward in bit array until next 1 is found. An even faster option is to read an integer from 
the bit array, fix it's offset and count trailing zeros. Traveling the bit array is recommended, since it is very simple to implement and is not noticably slower. See 
Table~\ref{endblockcall} for detailed times.

@@

@@

 - get actual results of rank vs select support size
 - take advantage of how data is stored

 - emphasis on size!

     RANK seems a lot faster, check algorithms!!!
TODO: try bitvector SD performance + size (select)

@@

The following experiments are run on <at the moment my work Lenovo Thinkpad T480s, TODO run on a server>.

@@

 - use array instead of vector<uint8 t> for bris?
- comparison to basic implementation + Bri09
  - compare with b=2,4,8

>>>>

\begin{table}
\centering
\caption{Size of support structure with SDSL-lite library}
\begin{tabular}{l||c c c c c c} 
Structure & TODO & get & multiple & size & calculations & here\\ 
\hline \hline 
$rank$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$rank_v5$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$select$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
\hline
%
\end{tabular}
\label{supportsize}
\end{table}

\section{DAC via rank}
Directly addressable codes in VB was first introduced by \citep{Bri09} in 2009. Their solution was to divide chunks in encoded bytes to separate arrays $A_1$ to 
$A_n$ and then use $rank_1$ to get direct access. For example, if an element needed 3 chunks when encoded, the least significant bits would go to $A_1$, second 
least significant bits to $A_2$ and the most significant bits would go to $A_3$. When fetching i-th element, getting the first chunk is just fetching $A_1$[i]. 
If the continuation bit is set at 0 (meaning there are more chunks to this element), getting the correct index for $A_2$ is calculated with $rank_1$($A_1$, i). 
Small values benefit from this reordering, because the first chunk does not need any calculation. This method causes a small overhead on the data, because each 
continuation bit array needs a support data structure for rank. 
A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8. In the actual implementation, different block size requires
extra calculation to get the block location from the byte array. (TODO: do i need to explain how to get the block location?) (TODO: does this need an example?)

\begin{figure}[ht]
\begin{algorithmic}
\State $i \gets $ wanted index
\State $A \gets $ block arrays
\State $B \gets $ continuation bit arrays
\State $level \gets 0$
\State $number \gets 0$
\While {$B[level][index] = 0$}
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number \mathbin{|} block$
\State $index \gets \Call{Rank}{B[level], index}$
\State $level \gets level + 1$
\EndWhile
\State $block \gets A[level][index]$
\State $number \gets number \mathbin{\ll} 8$
\State $number \gets number \mathbin{|} block$

\end{algorithmic}
\caption{Example pseudo code of DAC with rank by Brisaboa et al.} \label{bris_pseudo}
\end{figure}
<example of Bri09>

@@

The data in blocks should be written into memory as it is, so that when reading a large unsigned integer from the block byte array, the bits are already in correct order. 
This way two bit shift operations are enough to extract the desired value.  Assume original element size is 32 bits, block length is b and requested
element is at position x in the block array and it's compressed within k blocks. Starting byte s of element e is x * b div 8 where div is the integer division. Offset o
is the remainder of previous division, o = x*b mod 8. 32-bit value e is read from memory from s. Then we bit shift left 32-b*k-o to remove trailing bits and bit shift right
32-b*k to re-align bits.

Simple way to calculate block length k of i-th element is to get $select_1$(i+1) and subtract previously calculated start block index from it. This however causes a second 
$select_1$ query, which is costly. A much faster option is to iterate forward in bit array until next 1 is found. An even faster option is to read an integer from 
the bit array, fix it's offset and count trailing zeros. Traveling the bit array is preferred, since it is very simple to implement and is not noticably slower. See 
Table~\ref{endblockcall} for detailed times. See Figure~\ref{select_pseudo} for a DAC with $select$ and block size 8. With different block sizes, extra calculation to get 
the block location from the byte array is needed. In the example, $CalculateLength$ returns the length of the block and $ByteMask(k)$ returns a bit mask for k bytes. $number$
is a long unsigned integer with a length of multiple bytes.

@@

\begin{figure}[ht]
\begin{algorithmic}
\State $i \gets $ wanted index
\State $A \gets $ block byte array
\State $B \gets $ continuation bit array
\State $begin \gets \Call{Select}{index}$
\State $len \gets \Call{CalculateLength}{begin}$
\State $mask \gets \Call{ByteMask}{len}$
\State $number \gets A[begin]$ \Comment{read a larger number from the byte array}
\State $number \gets number \mathbin{\&} mask$

\end{algorithmic}
\caption{Pseudo code of DAC with select with block length 8} \label{select_pseudo}
\end{figure}

@@

 - get actual results of rank vs select support size

 - compare data structure size!

     RANK seems a lot faster on small arrays, check both algorithms!
TODO: try bitvector SD performance + size (select)

@@

The following experiments are run on <at the moment my work Lenovo Thinkpad T480s, TODO run on a server and write specs here>.

@@

 - do cachegrind
 - 
 - use array instead of vector<uint8 t> for bris?
- comparison to basic implementation + Bri09
  - compare with b=2,4,8
